Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-19 paper.tex
minors

We employ this approach in a swarm-like system setting where agents are tasked to cover phenomena detected in their environment. Over time, the agents have to converge over the each phenomenon in order to cover it appropriately as illustrated in \Cref{fig:simulations}.
\begin{figure}[t]
	\centering
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/start.png}
		\caption{Start}
		\label{fig:initial}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/after.png}
		\caption{After 50 steps}
		\label{fig:after}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/end.png}
		\caption{After 200 steps}
		\label{fig:end}
	\end{subfigure}
	\caption{Qualitative test results of the proposed approach. 
		The agents go towards the phenomenon and try to cover it without collapsing into a single point.
		the grey area represents the area covered by the agents (last figure on the right).}
	\label{fig:simulations}
\end{figure}

@@

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} compotations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a field computed in a previous round into a new field; 
\emph{ii)} the exchange of data with neighbouring fields (typically referred as \lstinline|nbr|), where received data is accessed by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
%We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation*}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\qquad
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation*}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation perceived at the time $t$: $o^v_t \in O$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022} if we set $f_v = o^v_t$.
\subsection{Motivation}
%The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
% \emph{manual design}, where the application is conceived and described through a programming language, 
% and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
%Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
As mentioned the above, our work falls between automatic and manual approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
%
Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesised through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

@@

In this section, we discuss the components involved in our proposed solution (i.e., architecture) and how these components interact with each other to bring the system to perform the collective behaviour (i.e dynamics). Finally, we will detail the learning algorithm designed and used to synthesise the policy.

@@

We generated 64 random scenarios for each type of experiment. Additionally, the placement of the agents were randomised as it has been done during training.
For the first type, considering a single static the phenomena is randomly placed in the environment. This is in contrast to the training where the phenomena was always placed in the same location.  
For the second type, we placed two distinct phenomena within the area. There location is kept constant on all 64 experiments. As the training only contained a single phenomena, this setup represents a challenge for the agents.
Finally, the third type contained a moving phenomena. In each scenario, the starting position as well as their direction of movement is randomly sampled from a uniform distribution. The movement is in a straight line with a constant speed of 5m/s within an unbounded environment. 

@@

\subsection{Baselines}
We compare our \ac{FIRL} approach against a baseline approaches where the \ac{DQN} utilises a \ac{MLP} as well as an approach only relying on \acp{GNN}, without additional field information. 
%
In all approaches, the underlying neural network (i.e., the \ac{MLP} and the \ac{GNN}) are trained %with the same parameters of the proposed approach 
with a single, stationary phenomenon.
%

@@

For all experiments, both \ac{GNN} versions demonstrate the capability to transfer the learned experience to the test phase whereas the MLP version fails to generalize. 
%
It is worth noting that, in the \emph{Zone Fixed} experiment, % (left graph in \Cref{fig:test}), 
 once the desired configuration is achieved in the static case, 
 the agents cease to move, maintaining the found configuration.
%
Interesting observations arise when we use scenarios different from the training phase. 
% Starting from a division into two zones, 
In the \emph{Two Zones} experiment,
 we notice that our approach using field-information finds a better configuration than the simple GNN counterpart. 
%
It exhibits both higher overall coverage and manages to divide the system into two equally covered parts. 
%
Indeed, observing the \Cref{fig:resCoverage}, we notice that the informed version maintains a balanced coverage between the two zones, 
 with a difference of less than 5\% between the two parts, maintaining a fair division of the phenomena.
%
In contrast, the uninformed version also maintains a fair division but with significantly different coverage between the two parts, 
 indicating a wrong placement among the agents in one of the zones.
%
This is a consequence of the uninformed version's inability to encode the necessary information to divide the agents into two zones, therefore it is not able to generalise.
%
Finally, the \emph{Moving} experiment emphasizes how the informed version generates a more robust policy for new scenarios. 
 Indeed, we observe that the our approach using \ac{FIRL} maintains higher coverage and a greater number of agents on the target phenomenon compared to the other two solutions.
%
The uninformed GNN version, however, 
 fails again to generalize its movement behaviour, as evidenced by the simulations where the agents, once reaching the target zone, stop moving due to tracking issues.

@@

\begin{figure*}
	\centering
  	\begin{subfigure}[b]{0.85\linewidth}
    \includegraphics[width=\linewidth]{imgs/inside-two-test}
    \caption{\emph{Two Zones} experiment: aggregated number of agent inside each phenomenon}
    \label{fig:inside-two-test}
  \end{subfigure}
  \begin{subfigure}[b]{0.85\linewidth}
    \includegraphics[width=\linewidth]{imgs/coverage-two-test}
    \caption{\emph{Two Zones} experiment: ratio of covered to uncovered area both phenomena}
    \label{fig:coverage-two-test}
  \end{subfigure} 
  \caption{Coverage of two zones using the different modes of the controller.}
  \label{fig:resCoverage}
\end{figure*}

>>>>

We employ this approach in a swarm-like system setting where agents are tasked to cover phenomena detected in their environment. Over time, the agents have to converge over each phenomenon in order to cover it appropriately as illustrated in \Cref{fig:simulations}.
\begin{figure}[t]
	\centering
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/start.png}
		\caption{Start}
		\label{fig:initial}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/after.png}
		\caption{After 50 steps}
		\label{fig:after}
	\end{subfigure}
	\begin{subfigure}[b]{0.32\linewidth}
		\includegraphics[width=\textwidth]{imgs/end.png}
		\caption{After 200 steps}
		\label{fig:end}
	\end{subfigure}
	\caption{Qualitative test results of the proposed approach. 
		The agents go towards the phenomenon and try to cover it without collapsing into a single point.
		the grey area represents the area covered by the agents (last figure on the right).}
	\label{fig:simulations}
\end{figure}

@@

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} compotations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a field computed in a previous round into a new field; 
\emph{ii)} the exchange of data with neighbouring fields (typically referred to as \lstinline|nbr|), where received data is accessed by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
%We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation*}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\qquad
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation*}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation perceived at the time $t$: $o^v_t \in O$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022} if we set $f_v = o^v_t$.
\subsection{Motivation}
%The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
% \emph{manual design}, where the application is conceived and described through a programming language, 
% and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
%Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
As mentioned above, our work falls between automatic and manual approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
%
Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesised through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

@@

In this section, we discuss the components involved in our proposed solution (i.e., architecture) and how these components interact with each other to bring the system to perform the collective behaviour (i.e., dynamics). Finally, we will detail the learning algorithm designed and used to synthesise the policy.

@@

We generated 64 random scenarios for each type of experiment. 
 Additionally, the placement of the agents was randomised as it has been done during training.
For the first type, consider a single static phenomenon randomly placed in the environment. 
This is in contrast to the training where the phenomena were always placed in the same location.  
For the second type, we placed two distinct phenomena within the area. 
Their location is kept constant in all 64 experiments. 
As the training only contained a single phenomenon, 
this setup represents a challenge for the agents.
Finally, the third type contained moving phenomena. 
In each scenario, the starting position as well as the direction of movement is randomly sampled from a uniform distribution. 
The movement is in a straight line with a constant speed of 5m/s within an unbounded environment. 

@@

\subsection{Baselines}
We compare our \ac{FIRL} approach against baseline approaches where the \ac{DQN} utilises a \ac{MLP} as well as an approach only relying on \acp{GNN}, without additional field information. 
%
In all approaches, the underlying neural network (i.e., the \ac{MLP} and the \ac{GNN}) are trained %with the same parameters of the proposed approach 
with a single, stationary phenomenon.
%

@@

For all experiments, both \ac{GNN} versions demonstrate the capability to transfer the learned experience to the test phase whereas the MLP version fails to generalize. 
%
It is worth noting that, in the \emph{Zone Fixed} experiment, % (left graph in \Cref{fig:test}), 
 once the desired configuration is achieved in the static case, 
 the agents cease to move, maintaining the found configuration.
%
Interesting observations arise when we use scenarios different from the training phase. 
% Starting from a division into two zones, 
In the \emph{Two Zones} experiment,
 we notice that our approach using field-information finds a better configuration than the simple GNN counterpart. 
%
It exhibits both higher overall coverage and manages to divide the system into two equally covered parts. 
%
Indeed, observing the \Cref{fig:resCoverage}, we notice that the informed version maintains a balanced coverage between the two zones, 
 with a difference of less than 5\% between the two parts, maintaining a fair division of the phenomena.
%
In contrast, the uninformed version also maintains a fair division but with significantly different coverage between the two parts, 
 indicating a wrong placement among the agents in one of the zones.
%
This is a consequence of the uninformed version's inability to encode the necessary information to divide the agents into two zones, therefore it is not able to generalise.
%
Finally, the \emph{Moving} experiment emphasizes how the informed version generates a more robust policy for new scenarios. 
 Indeed, we observe that our approach using \ac{FIRL} maintains higher coverage and a greater number of agents on the target phenomenon compared to the other two solutions.
%
The uninformed GNN version, however, 
 fails again to generalize its movement behaviour, as evidenced by the simulations where the agents, once reaching the target zone, stop moving due to tracking issues.

@@

\begin{figure*}
	\centering
  	\begin{subfigure}[b]{0.85\linewidth}
    \includegraphics[width=\linewidth]{imgs/inside-two-test}
    \caption{\emph{Two Zones} experiment: aggregated number of agent inside each phenomenon}
    \label{fig:inside-two-test}
  \end{subfigure}
  \begin{subfigure}[b]{0.85\linewidth}
    \includegraphics[width=\linewidth]{imgs/coverage-two-test}
    \caption{\emph{Two Zones} experiment: ratio of covered to the uncovered area both phenomena}
    \label{fig:coverage-two-test}
  \end{subfigure} 
  \caption{Coverage of two zones using the different modes of the controller.}
  \label{fig:resCoverage}
\end{figure*}
